Graded Call-By-Push-Value

### 2020-07-07 popl21/popl21.tex
[ popl21 ] old coeffect grading (<> instead of < r >)

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

@@

\newcommand{\denp}{\den[+]}
\newcommand{\denn}{\den[-]}
\newcommand{\Denn}[2]{\denn{#1}_{#2}}

\newcommand{\tEff}{\ensuremath{\mathsf{Eff}}}
\newcommand{\Eff}{\ensuremath{\mathsf{Eff}}}
\newcommand{\bu}{\ensuremath{\bullet}}
\newcommand{\Ge}{\ensuremath{\varepsilon}}
\newcommand{\Ga}{\ensuremath{\alpha}}
\newcommand{\Gl}{\ensuremath{\lambda}}
\newcommand{\Gr}{\ensuremath{\rho}}
\newcommand{\GG}{\ensuremath{\Gamma}}
\newcommand{\GS}{\ensuremath{\Sigma}}
\newcommand{\R}{\mathsf{R}}
\newcommand{\I}{\mathsf{I}}
\newcommand{\LL}{\mathcal{L}}
\newcommand{\W}{\mathsf{W}}
\newcommand{\M}{\mathsf{M}}
\newcommand{\wleq}{\sqsubseteq}
\newcommand{\winf}{\sqcap}
\newcommand{\wsup}{\sqcup}
%\newcommand{\winf}{\bigsqcap}
\newcommand{\emp}{0_{\W}}
% \newcommand{\emp}{\varepsilon}
% \newcommand{\emp}{\emptyset}

\newcommand{\CC}{\ensuremath{\mathcal{C}}}
\newcommand{\Hom}[2]{\CC\,(#1,\,#2)}
\newcommand{\SET}{\mathsf{SET}}
\newcommand{\CPO}{\mathsf{CPO}}
\newcommand{\T}[1][]{\mathsf{T}_{#1}}
\newcommand{\D}[1][]{\mathsf{D}_{#1}}
\newcommand{\trun}{\mathsf{run}}
\newcommand{\run}[1][]{\mathsf{run}_{#1}}
\newcommand{\texpose}{\mathsf{expose}}
\newcommand{\expose}[1][]{\texpose_{#1}}
\newcommand{\tA}{\mathsf{A}}
\newcommand{\tB}{\mathsf{B}}
\newcommand{\A}{\mathsf{A}}
\newcommand{\B}{\mathsf{B}}
\newcommand{\tExc}{\mathsf{Exc}}
\newcommand{\Exc}{\mathsf{Exc}}
\newcommand{\String}{\mathsf{String}}
\newcommand{\StringLe}[1]{\mathsf{String}{\leq}\,#1}
\newcommand{\tfmap}{\mathsf{fmap}}
\newcommand{\fmap}[2][]{\tfmap_{#1}\,#2}
\newcommand{\treturn}{\mathsf{return}}
\newcommand{\tjoin}{\mathsf{join}}
\newcommand{\join}{\mathsf{join}}
\newcommand{\append}{\ensuremath{\mathbin{\mathtt{++}}}}
\newcommand{\tcast}{\mathsf{cast}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\tid}{\mathsf{id}}
\newcommand{\teval}{\mathsf{eval}}
\newcommand{\tstrength}{\mathsf{strength}}
\newcommand{\strengthl}{\tstrength^{\tl}}
\newcommand{\strengthr}{\tstrength^{\tr}}
\newcommand{\tl}{\mathsf{l}}
\newcommand{\tr}{\mathsf{r}}
\newcommand{\tcurry}{\mathsf{curry}}
\newcommand{\comp}{\circ}
\newcommand{\tweaken}{\mathsf{weaken}}
\newcommand{\tcontract}{\mathsf{contract}}

% Types
\newcommand{\tTy}{\mathsf{Ty}}
\newcommand{\Ty}[1]{\tTy^{#1}}
\newcommand{\PTy}{\Ty+}
\newcommand{\NTy}{\Ty-}
\newcommand{\extr}[2]{#1.#2}
\newcommand{\ext}[3]{#1.#2{:}#3}
\newcommand{\qext}[5]{#1#2.#3#4{:}#5}
\newcommand{\gqext}[5]{\graybox{#1}#2.\graybox{#3}#4{:}#5}

@@

\begin{figure}[htbp]
\flushleft
\ruler{}
Types.
\[
\begin{array}{lllrl@{\qquad}l}
\PTy & \ni & P
  & ::= & \graybox{\thunkty e N}
     \mid o \mid \sumty i I P \mid \tupty i I P
  & \mbox{Value types (positive types)} \\
\NTy & \ni & N
  & ::= & \diamond P
     \mid P \To N \mid \recty i I N
  & \mbox{Computation types (negative types)} \\
\end{array}
\]
\dashruler{}
Terms.
\[
\begin{array}{lllrl@{\qquad}l}
\PTm & \ni & v,w
  & ::= & x
     \mid \thunk t
     \mid \inj i v
     \mid \ptup{\bar v}
  & \mbox{Values (positive terms)} \\
\NTm & \ni & t,u
  & ::= &
         \vlet x v t
  & \mbox{Computations (negative terms):}
\\ &&& \mid &
        \force v
    \mid \case v {\overline{x.\,t}}
    \mid \splits v {\bar x} t
  & \mbox{value eliminations}
\\ &&& \mid &
         \ret v        \mid \bind x u t
  & \mbox{monad operations}
\\ &&& \mid &
         \lam x t      \mid \app t v
  & \mbox{functions}
\\ &&& \mid &
         \trecord\{\overline{i:t}\} % \ntup{\bar t}
       \mid \proj i t
  & \mbox{lazy tuples (records)}
\end{array}
\]
\dashruler{}
Value typing \fbox{$\ValTy \Gamma v P$}\,.
\begin{gather*}
 \nru{\rvar}
     {x{:}P \in \Gamma
    }{\ValTy \Gamma x P}
\qquad
 \nru{\rintro\Box}
     {\CompTy \Gamma t {\graybox e} N
    }{\ValTy \Gamma {\thunk t} {\thunkty {\graybox e} N}}
\\[1.5ex]
 \nru{\rintro\GS}
     {\ValTy \Gamma v {P_i}
    }{\ValTy \Gamma {\inj i v} {\sumt I P}}
\qquad
 \nru{\rintro\otimes}
     {\forall i{:}I, \ \ValTy \Gamma {v_i} {P_i}
    }{\ValTy \Gamma {\ptup v} {\tupt I P}}
    % }{\ValTy \Gamma {\ptup{\overline{v_i}^{i:I}}} {\tupt I P}}
%    }{\ValTy \Gamma {\ptup (v_i)_{i:I}} {\tupt I P}}
\end{gather*}
\dashruler{}
Computation typing \fbox{$\CompTy \Gamma t {\graybox{e}} N$}\,.
\begin{gather*}
 \nru{\rlet}
     {\ValTy \Gamma v P \qquad
      \CompTy {\ext \Gamma x P} t e N
    }{\CompTy \Gamma {\vlet x v t} e N}
\qquad
 \nru{\relim\Box}
     {\ValTy \Gamma v {\thunkty {\graybox e} N}
    }{\CompTy \Gamma {\force v} {\graybox e} N}
\\[1.5ex]
 \nru{\relim\GS}
     {\ValTy \Gamma v {\sumt I P} \qquad
      \forall i{:}I,\ \CompTy {\ext \Gamma {x_i} {P_i}} {t_i} e N
    }{\CompTy \Gamma {\caser v {\{x_i.\, t_i\}_{i:I}}} e N}
\qquad
 \nru{\relim\otimes}
     {\ValTy \Gamma v {\tupt I P} \qquad
      \CompTy {\extr \Gamma {\overline{x_i{:}P_i}^{i:I}}} t e N
    }{\CompTy \Gamma {\splits v {\bar x} t} e N}
\\[1.5ex]
 \nru{\rintro\diamond}
     {\ValTy \Gamma v P
    }{\CompTy \Gamma {\ret v} {\graybox \Ge} {\diamond P}}
\qquad
 \nru{\relim\diamond}
     {\CompTy \Gamma u {\graybox{e_1}} {\diamond P} \qquad
      \CompTy {\ext \Gamma x P} t {\graybox{e_2}} N
    }{\CompTy \Gamma {\bind x u t} {\graybox{e_1 \bu e_2}} N}
\\[1.5ex]
 \nru{\rintro\To}
     {\CompTy {\ext \Gamma x P} t e N
    }{\CompTy \Gamma {\lam x t} e {P \To N}}
\qquad
 \nru{\relim\To}
     {\CompTy \Gamma t e {P \To N} \qquad
      \ValTy \Gamma v P
    }{\CompTy \Gamma {\app t v} e N}
\\[1.5ex]
 \nru{\rintro\Pi}
     {\forall i{:}I,\ \CompTy \Gamma {t_i} e {N_i}
    }{\CompTy \Gamma {\trecord {\{i : t_i\}_{i:I}}} e {\rect I N}}
    % }{\CompTy \Gamma {\ntup{\overline{t_i}^{i:I}}} e {\rect I N}}
\qquad
 \nru{\relim\Pi}
     {\CompTy \Gamma t e {\rect I N}
    }{\CompTy \Gamma {\proj i t} e  {N_i}}
\\[1.5ex]
\graybox{
 \nrux{\rsub}
      {\CompTy \Gamma t e N}
      {\CompTy \Gamma t {e'} N}
      {e \leq e'}
}
\end{gather*}
\rule{\textwidth}{0.2pt}
  \caption{Effect-graded call-by-push-value.}
  \label{fig:cbpv}
\end{figure}

@@

\newcommand{\lolli}{\multimap}
\newcommand{\qfun}[3]{#1#2 \lolli #3}
\newcommand{\gqfun}[3]{\graybox{#1#2 \lolli #3}}
\renewcommand{\CompTy}[3]{#1 \vdash #2 : #3}
\newcommand{\nbox}[1]{}
\newcommand{\tzip}{\mathsf{zip}}
\newcommand{\textract}{\mathsf{extract}}
\newcommand{\extract}[1][]{\textract_{#1}}
\newcommand{\tduplicate}{\mathsf{duplicate}}
\newcommand{\duplicate}[1][]{\tduplicate_{#1}}
\newcommand{\tcobind}{\mathsf{cobind}}
\newcommand{\tbind}{\mathsf{bind}}
\newcommand{\tget}{\mathsf{get}}
\newcommand{\tmodify}{\mathsf{modify}}
\newcommand{\thead}{\mathsf{head}}
\newcommand{\ttail}{\mathsf{tail}}
\newcommand{\suc}{(\_{+}1)}
\newcommand{\Float}{\mathsf{Float}}

@@

% Let us consider the \emph{context comonad} $\D\,B = S \times B$ that
% gives us access to a value in $S$.
% \footnote{\url{https://bartoszmilewski.com/2017/01/02/comonads/}}
% Besides functoriality, a comonad
% has the natural transformations $\textract$ and $\duplicate$, dual to
% $\treturn$ and $\tjoin$.  The implementation of the context comonad is
% as follows:
% \[
% \begin{array}{lcl}
%   \textract & : & \D\,B \to B \\
%   \textract\,(s,b) & = & b
% \\[1.5ex]
%   \tduplicate & : & \D,\B \to \D\,(\D\,B) \\
%   \tduplicate\,(s,b) & = & (s, (s, b))
% \\[1.5ex]
%   \tget & : & \D\,B \to S \\
%   \tget\,(s,b) & = & s
% \\[1.5ex]
%   \tmodify
% \end{array}
% \]
% The context comonad is not monoidal.

Let us consider the \emph{stream comonad}%
\footnote{\url{https://bartoszmilewski.com/2017/01/02/comonads/}}
$\D\,B = \bN \to B$,
with $\thead\,s = s\,0$ and $\ttail\,s = s \comp \suc$.
The stream comonad lets us work in a setting where values are not
single data points, but streams of data.
Besides functoriality, a
comonad has the natural transformations $\textract$ and $\duplicate$,
dual to $\treturn$ and $\tjoin$.  The implementation of the stream
comonad in $\SET$ is as follows:
\[
\begin{array}{lcl}
  \textract & : & \D\,B \to B \\
  \textract\,s & = & \thead\,s
\\[1.5ex]
  \tduplicate & : & \D\,B \to \D\,(\D\,B) \\
  \thead\,(\tduplicate\,s) & = & s \\
  \ttail\,(\tduplicate\,s) & = & \tduplicate\,(\ttail\,s) \\
\end{array}
\]
The comonad laws are a simple dualization of the monad laws:
\[
\xymatrix@C=10ex{
\D\,B  \ar@{=}[dr]
  & \D\,(\D\,B) \ar[l]_{\textract} \ar[r]^{\tduplicate}
  & \D\,(\D\,(\D\,B))
\\
  & \D\,B \ar[u]_{\tduplicate} \ar[r]^{\tduplicate} \ar@{=}[dr]
  & \D\,(\D\,B) \ar[u]_{\tfmap\,\tduplicate} \ar[d]^{\tfmap\,\textract}
\\
  &
  & \D\,B
}
\]

@@

If type constructor $\Box$ is interpreted as a monoidal comonad $\D$,
then all positive types $P$ can be intepreted as $\D$-coalgebras
$A = \den P$, \ie, implement a morphism $\expose[A] : A \to \D\,A$
satisfying the laws of a comonad coalgebra:
\[
\xymatrix@C=10ex{
A  \ar@{=}[dr]
  & \D\,A \ar[l]_{\textract} \ar[r]^{\tduplicate}
  & \D\,(\D\,A)
\\
  & A \ar[u]_{\texpose} \ar[r]^{\texpose}
  & \D\,A \ar[u]_{\tfmap\,\texpose}
}
\]
The implementation of $\expose{\den P}$ proceeds by induction on $P$.
\[
\begin{array}{lcl}
  \expose{\den{\Box N}} & : & {\D\,\den N} \to {\D\,(\D\,\den N)} \\
  \expose{\den{\Box N}} & = & \duplicate
\\[1ex]
  \expose{\den{\tupt I P}} & : & \tupt {i:I} {\den{P_i}} \to \D\,(\tupt {i:I} {\den{P_i}}) \\
  \expose{\den{\tupt I P}} & = & \tzip \comp \tupt {i:I} \expose{\den{P_i}}
\\[1ex]
  \expose{\den{\sumt I P}} & : & \sumt {i:I} {\den{P_i}} \to \D\,(\sumt {i:I} {\den{P_i}}) \\
  \expose{\den{\sumt I P}} & = & [\tfmap\,\iota_i \comp \expose{\den{P_i}}]_{i:I} \\
\end{array}
\]
Positive base types $o \in \PTy$ are required to be $\D$-coalgebras.
For instance, elements of a base type like
$\Float \in \PTy$ could represent streams of floating point numbers,
\eg, continuous measurements from a sensor.

@@

Graded comonads have been utilized to give semantics to
context-dependent computation \citep{orchard:icfp14}.  For grading,
loccit uses a \emph{resource algebra} in form of a pre-ordered
semiring
$(\R,{+},0,{\cdot},1,\leq)$.
% $(\R,\_{+}\_,0,\_{\cdot}\_,1,\leq)$.
%
The semantics is \emph{resource aware} and thus not constructed in a
cartesian-closed category but in a symmetric monoidal closed category
$(\CC,\otimes,\I,\lolli)$.
%
As a running example, we use the semiring $\R = \Pot\bN \setminus \emptyset$ of
\emph{multiplicities} ordered by $\supseteq$ with pointwise sum and
product, \eg, $r + s = \{ n + m \mid n \in r, m \in s \}$.
Subsemirings of $\R$ have been used for quantitative typing
\cite{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics08}.
The order expresses precision of the quantities, \eg,
$\{1\} \geq \{0,1\} \geq \bN$ states that linear use of a variable is more informative than affine use than unrestricted use.

@@

@@

@@

  \label{fig:cbpv}

@@

>>>>

\nonstopmode
%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

@@

\newcommand{\denp}{\den[+]}
\newcommand{\denn}{\den[-]}
\newcommand{\Denn}[2]{\denn{#1}_{#2}}
\newcommand{\Den}[2]{\den{#1}_{#2}}

\newcommand{\tEff}{\ensuremath{\mathsf{Eff}}}
\newcommand{\Eff}{\ensuremath{\mathsf{Eff}}}
\newcommand{\bu}{\ensuremath{\bullet}}
\newcommand{\Ge}{\ensuremath{\varepsilon}}
\newcommand{\Ga}{\ensuremath{\alpha}}
\newcommand{\Gd}{\ensuremath{\delta}}
\newcommand{\Gl}{\ensuremath{\lambda}}
\newcommand{\Gr}{\ensuremath{\rho}}
\newcommand{\GG}{\ensuremath{\Gamma}}
\newcommand{\GS}{\ensuremath{\Sigma}}
\newcommand{\R}{\mathsf{R}}
\newcommand{\I}{\mathsf{I}}
\newcommand{\LL}{\mathcal{L}}
\newcommand{\W}{\mathsf{W}}
\newcommand{\M}{\mathsf{M}}
\newcommand{\wleq}{\sqsubseteq}
\newcommand{\winf}{\sqcap}
\newcommand{\wsup}{\sqcup}
%\newcommand{\winf}{\bigsqcap}
\newcommand{\emp}{0_{\W}}
% \newcommand{\emp}{\varepsilon}
% \newcommand{\emp}{\emptyset}

\newcommand{\CC}{\ensuremath{\mathcal{C}}}
\newcommand{\Hom}[2]{\CC\,(#1,\,#2)}
\newcommand{\SET}{\mathsf{SET}}
\newcommand{\CPO}{\mathsf{CPO}}
\newcommand{\T}[1][]{\mathsf{T}_{#1}}
\newcommand{\D}[1][]{\mathsf{D}_{#1}}
\newcommand{\trun}{\mathsf{run}}
\newcommand{\run}[1][]{\mathsf{run}_{#1}}
\newcommand{\texpose}{\mathsf{expose}}
\newcommand{\expose}[1][]{\texpose_{#1}}
\newcommand{\tA}{\mathsf{A}}
\newcommand{\tB}{\mathsf{B}}
\newcommand{\A}{\mathsf{A}}
\newcommand{\B}{\mathsf{B}}
\newcommand{\tExc}{\mathsf{Exc}}
\newcommand{\Exc}{\mathsf{Exc}}
\newcommand{\String}{\mathsf{String}}
\newcommand{\StringLe}[1]{\mathsf{String}{\leq}\,#1}
\newcommand{\tfmap}{\mathsf{fmap}}
\newcommand{\fmap}[2][]{\tfmap_{#1}\,#2}
\newcommand{\treturn}{\mathsf{return}}
\newcommand{\tjoin}{\mathsf{join}}
\newcommand{\join}{\mathsf{join}}
\newcommand{\append}{\ensuremath{\mathbin{\mathtt{++}}}}
\newcommand{\tcast}{\mathsf{cast}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\tid}{\mathsf{id}}
\newcommand{\teval}{\mathsf{eval}}
\newcommand{\tstrength}{\mathsf{strength}}
\newcommand{\strengthl}{\tstrength^{\tl}}
\newcommand{\strengthr}{\tstrength^{\tr}}
\newcommand{\tl}{\mathsf{l}}
\newcommand{\tr}{\mathsf{r}}
\newcommand{\tcurry}{\mathsf{curry}}
\newcommand{\comp}{\circ}
\newcommand{\tweaken}{\mathsf{weaken}}
\newcommand{\tdrop}[1][]{\mathsf{drop}_{#1}}
%\newcommand{\drop}[1][]{\tdrop_{#1}} % Name clash
\newcommand{\tcontract}{\mathsf{contract}}
\newcommand{\contract}[1][]{\tcontract_{#1}}

% Types
\newcommand{\tTy}{\mathsf{Ty}}
\newcommand{\Ty}[1]{\tTy^{#1}}
\newcommand{\PTy}{\Ty+}
\newcommand{\NTy}{\Ty-}
\newcommand{\Cxt}{\mathsf{Cxt}}
\newcommand{\RCxt}{\mathsf{RCxt}}
\newcommand{\cempty}{\emptyset}
\newcommand{\extr}[2]{#1.#2}
\newcommand{\ext}[3]{#1.#2{:}#3}
\newcommand{\qext}[5]{#1#2.#3#4{:}#5}
\newcommand{\gqext}[5]{\graybox{#1}#2.\graybox{#3}#4{:}#5}

@@

\begin{figure}[htbp]
\flushleft
\ruler{}
Types.
\[
\begin{array}{lllrl@{\qquad}l}
\PTy & \ni & P
  & ::= & \graybox{\thunkty e N}
     \mid o \mid \sumty i I P \mid \tupty i I P
  & \mbox{Value types (positive types)} \\
\NTy & \ni & N
  & ::= & \diamond P
     \mid P \To N \mid \recty i I N
  & \mbox{Computation types (negative types)} \\
\Cxt & \ni & \Gamma
  & ::= & \cempty \mid \ext \Gamma x P
  & \mbox{Typing context}
\\
\end{array}
\]
\dashruler{}
Terms.
\[
\begin{array}{lllrl@{\qquad}l}
\PTm & \ni & v,w
  & ::= & x
     \mid \thunk t
     \mid \inj i v
     \mid \ptup{\bar v}
  & \mbox{Values (positive terms)} \\
\NTm & \ni & t,u
  & ::= &
         \vlet x v t
  & \mbox{Computations (negative terms):}
\\ &&& \mid &
        \force v
    \mid \case v {\overline{x.\,t}}
    \mid \splits v {\bar x} t
  & \mbox{value eliminations}
\\ &&& \mid &
         \ret v        \mid \bind x u t
  & \mbox{monad operations}
\\ &&& \mid &
         \lam x t      \mid \app t v
  & \mbox{functions}
\\ &&& \mid &
         \trecord\{\overline{i:t}\} % \ntup{\bar t}
       \mid \proj i t
  & \mbox{lazy tuples (records)}
\end{array}
\]
\dashruler{}
Value typing \fbox{$\ValTy \Gamma v P$}\,.
\begin{gather*}
 \nru{\rvar}
     {x{:}P \in \Gamma
    }{\ValTy \Gamma x P}
\qquad
 \nru{\rintro\Box}
     {\CompTy \Gamma t {\graybox e} N
    }{\ValTy \Gamma {\thunk t} {\thunkty {\graybox e} N}}
\\[1.5ex]
 \nru{\rintro\GS}
     {\ValTy \Gamma v {P_i}
    }{\ValTy \Gamma {\inj i v} {\sumt I P}}
\qquad
 \nru{\rintro\otimes}
     {\forall i{:}I, \ \ValTy \Gamma {v_i} {P_i}
    }{\ValTy \Gamma {\ptup v} {\tupt I P}}
    % }{\ValTy \Gamma {\ptup{\overline{v_i}^{i:I}}} {\tupt I P}}
%    }{\ValTy \Gamma {\ptup (v_i)_{i:I}} {\tupt I P}}
\end{gather*}
\dashruler{}
Computation typing \fbox{$\CompTy \Gamma t {\graybox{e}} N$}\,.
\begin{gather*}
 \nru{\rlet}
     {\ValTy \Gamma v P \qquad
      \CompTy {\ext \Gamma x P} t e N
    }{\CompTy \Gamma {\vlet x v t} e N}
\qquad
 \nru{\relim\Box}
     {\ValTy \Gamma v {\thunkty {\graybox e} N}
    }{\CompTy \Gamma {\force v} {\graybox e} N}
\\[1.5ex]
 \nru{\relim\GS}
     {\ValTy \Gamma v {\sumt I P} \qquad
      \forall i{:}I,\ \CompTy {\ext \Gamma {x_i} {P_i}} {t_i} e N
    }{\CompTy \Gamma {\caser v {\{x_i.\, t_i\}_{i:I}}} e N}
\qquad
 \nru{\relim\otimes}
     {\ValTy \Gamma v {\tupt I P} \qquad
      \CompTy {\extr \Gamma {\overline{x_i{:}P_i}^{i:I}}} t e N
    }{\CompTy \Gamma {\splits v {\bar x} t} e N}
\\[1.5ex]
 \nru{\rintro\diamond}
     {\ValTy \Gamma v P
    }{\CompTy \Gamma {\ret v} {\graybox \Ge} {\diamond P}}
\qquad
 \nru{\relim\diamond}
     {\CompTy \Gamma u {\graybox{e_1}} {\diamond P} \qquad
      \CompTy {\ext \Gamma x P} t {\graybox{e_2}} N
    }{\CompTy \Gamma {\bind x u t} {\graybox{e_1 \bu e_2}} N}
\\[1.5ex]
 \nru{\rintro\To}
     {\CompTy {\ext \Gamma x P} t e N
    }{\CompTy \Gamma {\lam x t} e {P \To N}}
\qquad
 \nru{\relim\To}
     {\CompTy \Gamma t e {P \To N} \qquad
      \ValTy \Gamma v P
    }{\CompTy \Gamma {\app t v} e N}
\\[1.5ex]
 \nru{\rintro\Pi}
     {\forall i{:}I,\ \CompTy \Gamma {t_i} e {N_i}
    }{\CompTy \Gamma {\trecord {\{i : t_i\}_{i:I}}} e {\rect I N}}
    % }{\CompTy \Gamma {\ntup{\overline{t_i}^{i:I}}} e {\rect I N}}
\qquad
 \nru{\relim\Pi}
     {\CompTy \Gamma t e {\rect I N}
    }{\CompTy \Gamma {\proj i t} e  {N_i}}
\\[1.5ex]
\graybox{
 \nrux{\rsub}
      {\CompTy \Gamma t e N}
      {\CompTy \Gamma t {e'} N}
      {e \leq e'}
}
\end{gather*}
\rule{\textwidth}{0.2pt}
  \caption{Effect-graded call-by-push-value.}
  \label{fig:cbpv}
\end{figure}

@@

\newcommand{\lolli}{\multimap}
\newcommand{\qfun}[3]{#1#2 \lolli #3}
\newcommand{\gqfun}[3]{\graybox{#1#2 \lolli #3}}
\renewcommand{\CompTy}[3]{#1 \vdash #2 : #3}
\newcommand{\nbox}[1]{}
\newcommand{\tzip}{\mathsf{zip}}
\newcommand{\textract}{\mathsf{extract}}
\newcommand{\extract}[1][]{\textract_{#1}}
\newcommand{\tdisplay}{\mathsf{display}}
\newcommand{\display}[1][]{\tdisplay_{#1}}
% \newcommand{\tduplicate}{\mathsf{display}}
% \newcommand{\duplicate}[1][]{\tdisplay_{#1}}
\newcommand{\tduplicate}{\mathsf{duplicate}}
\newcommand{\duplicate}[1][]{\tduplicate_{#1}}
\newcommand{\tcobind}{\mathsf{cobind}}
\newcommand{\tbind}{\mathsf{bind}}
\newcommand{\tget}{\mathsf{get}}
\newcommand{\tmodify}{\mathsf{modify}}
\newcommand{\thead}{\mathsf{head}}
\newcommand{\ttail}{\mathsf{tail}}
\newcommand{\suc}{(\_{+}1)}
\newcommand{\Float}{\mathsf{Float}}
\newcommand{\tiso}{\mathsf{iso}}
\newcommand{\isot}{\tiso^\otimes}

@@

% Let us consider the \emph{context comonad} $\D\,B = S \times B$ that
% gives us access to a value in $S$.
% \footnote{\url{https://bartoszmilewski.com/2017/01/02/comonads/}}
% Besides functoriality, a comonad
% has the natural transformations $\textract$ and $\display$, dual to
% $\treturn$ and $\tjoin$.  The implementation of the context comonad is
% as follows:
% \[
% \begin{array}{lcl}
%   \textract & : & \D\,B \to B \\
%   \textract\,(s,b) & = & b
% \\[1.5ex]
%   \tdisplay & : & \D,\B \to \D\,(\D\,B) \\
%   \tdisplay\,(s,b) & = & (s, (s, b))
% \\[1.5ex]
%   \tget & : & \D\,B \to S \\
%   \tget\,(s,b) & = & s
% \\[1.5ex]
%   \tmodify
% \end{array}
% \]
% The context comonad is not monoidal.

Let us consider the \emph{stream comonad}%
\footnote{\url{https://bartoszmilewski.com/2017/01/02/comonads/}}
$\D\,B = \bN \to B$,
with $\thead\,s = s\,0$ and $\ttail\,s = s \comp \suc$.
The stream comonad lets us work in a setting where values are not
single data points, but streams of data.
Besides functoriality, a
comonad has the natural transformations $\textract$ and $\display$
(in category theory called $\Ge$ and $\Gd$),
dual to $\treturn$ and $\tjoin$.  The implementation of the stream
comonad in $\SET$ is as follows:
\[
\begin{array}{lcl}
  \textract & : & \D\,B \to B \\
  \textract\,s & = & \thead\,s
\\[1.5ex]
  \tdisplay & : & \D\,B \to \D\,(\D\,B) \\
  \thead\,(\tdisplay\,s) & = & s \\
  \ttail\,(\tdisplay\,s) & = & \tdisplay\,(\ttail\,s) \\
\end{array}
\]
The generic operations of a comonad allow us to \emph{extract} the
value wrapped in a comonadic structure and to \emph{display} another
layer of that structure.

The comonad laws are a simple dualization of the monad laws:
\[
\xymatrix@C=10ex{
\D\,B  \ar@{=}[dr]
  & \D\,(\D\,B) \ar[l]_{\textract} \ar[r]^{\tdisplay}
  & \D\,(\D\,(\D\,B))
\\
  & \D\,B \ar[u]_{\tdisplay} \ar[r]^{\tdisplay} \ar@{=}[dr]
  & \D\,(\D\,B) \ar[u]_{\tfmap\,\tdisplay} \ar[d]^{\tfmap\,\textract}
\\
  &
  & \D\,B
}
\]

@@

If type constructor $\Box$ is interpreted as a monoidal comonad $\D$,
then all positive types $P$ can be intepreted as $\D$-coalgebras
$A = \den P$, \ie, implement a morphism $\expose[A] : A \to \D\,A$
satisfying the laws of a comonad coalgebra:
\[
\xymatrix@C=10ex{
A  \ar@{=}[dr]
  & \D\,A \ar[l]_{\textract} \ar[r]^{\tdisplay}
  & \D\,(\D\,A)
\\
  & A \ar[u]_{\texpose} \ar[r]^{\texpose}
  & \D\,A \ar[u]_{\tfmap\,\texpose}
}
\]
The implementation of $\expose[\den P]$ proceeds by induction on $P$.
\[
\begin{array}{lcl}
  \expose[\den{\Box N}] & : & {\D\,\den N} \to {\D\,(\D\,\den N)} \\
  \expose[\den{\Box N}] & = & \display
\\[1ex]
  \expose[\den{\tupt I P}] & : & \tupt {i:I} {\den{P_i}} \to \D\,(\tupt {i:I} {\den{P_i}}) \\
  \expose[\den{\tupt I P}] & = & \tzip \comp \tupt {i:I} \expose[\den{P_i}]
\\[1ex]
  \expose[\den{\sumt I P}] & : & \sumt {i:I} {\den{P_i}} \to \D\,(\sumt {i:I} {\den{P_i}}) \\
  \expose[\den{\sumt I P}] & = & \left[\tfmap\,\iota_i \comp \expose[\den{P_i}]\right]_{i:I} \\
\end{array}
\]
Positive base types $o \in \PTy$ are required to be $\D$-coalgebras.
For instance, elements of a base type like
$\Float \in \PTy$ could represent streams of floating point numbers,
\eg, continuous measurements from a sensor.

@@

Graded comonads have been utilized to give semantics to
context-dependent computation \citep{orchard:icfp14}.  For grading,
loccit uses a \emph{resource algebra} in form of a pre-ordered
semiring
$(\R,{+},0,{\cdot},1,\leq)$.
% $(\R,\_{+}\_,0,\_{\cdot}\_,1,\leq)$.
%
The semantics is \emph{resource aware} and thus not constructed in a
cartesian-closed category but in a symmetric monoidal closed category
$(\CC,\otimes,\I,\lolli)$.  We overload the operators on
exponentials, $\Lambda : \Hom {C \otimes A} B \to \Hom C {A \lolli B}$
and $\teval : \Hom{(A \lolli B) \otimes A} B$.  The tensor product
$\otimes$ is a bifunctor and it is customary to overload $\otimes$ for
the functorial action
$f_1 \otimes f_2 : \Hom{A_1 \otimes A_2}{B_1 \otimes B_2}$ where
$f_i : \Hom{A_i}{B_i}$.  We shall also use $I$-ary products
$\tupt {i:I} {A_i}$ and their functorial action $\tupt {i:I} {f_i}$.
Further, the symmetric monoidal structure is usually witnessed by
isomorphisms for left $\lambda_A : \Hom {\I \otimes A} A$ and right
unit $\rho_A : \Hom {A \otimes I} A$, associativity
$\Ga_{A,B,C} : \Hom{A \otimes (B \otimes C)} {(A \otimes B) \otimes
  C}$
and swap $\sigma_{A,B} : \Hom {A \otimes B} {B \otimes A}$.  We shall
summarize combinations of these isomorphisms under the name
$\isot : \Hom {\tupt I A} {\tupt J B}$ when the multisets
$\{A_i\}_{i:I}$ and $\{B_j\}_{j:J}$ coincide modulo addition and
deletion of $\I$s.

As a running example for resource accounting,
we use the semiring $\R = \Pot\bN \setminus \emptyset$ of
\emph{multiplicities} ordered by $\supseteq$ with pointwise sum and
product, \eg, $r + s = \{ n + m \mid n \in r, m \in s \}$.
Subsemirings of $\R$ have been used for quantitative typing
\cite{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics08}.
The order expresses precision of the quantities, \eg,
$\{1\} \geq \{0,1\} \geq \bN$ states that linear use of a variable is more informative than affine use than unrestricted use.

@@

\Cxt & \ni & \Gamma
  & ::= & \cempty \mid \ext \Gamma x P \mid \graybox{\ext \Gamma y N}
  & \nbox{Typing context}
\\
\R & \ni & \graybox{r,s}
   & ::= & 0 \mid 1 \mid r + s \mid rs
   & \mbox{Resource qualifiers}
\\
\RCxt & \ni & \graybox{\gamma,\delta}
  & ::= & \cempty \mid \ext \gamma x r \mid \ext \gamma y r
  & \mbox{Resource context}

@@

%\\[1.5ex]

@@

  \label{fig:coeff-cbpv}

@@

\subsection{Coeffect-graded CBPV: typing and semantics}

We now have the mathematical structures in place to define a
coeffect-graded variant of CBPV (see Fig.~\ref{coeff-cbpv}).
The difference to pure CBPV is laid out in \emph{gray boxes}.
